Smart Spreading Factor Assignment for LoRaWANs

### 2019-01-14 paper.tex
shortened

\begin{abstract}
Low power wide area network (LPWAN) technologies offer affordable connectivity to massive number of low-power devices distributed over very large geographical areas. Focus of this work is one of the most promising LPWAN technologies: LoRa. LoRa offers long range communication and strong resilience to interference by proprietary modulation technique based on Chirp Spread Spectrum (CSS). LoRa modulation trades data rate for sensitivity and communication range by spreading symbols within a fixed channel bandwidth. Collisions in LoRaWAN networks are strongly related with spreading factor (SF) assignment of nodes which indeed effects network performance. In this work, a simulation environment to study different SF assignment schemes is implemented. Furthermore, a novel SF assignment strategy is proposed which utilizes a machine learning technique for optimization of SF assignment. Finally, promising simulation results for proposed machine learning SF assignment technique is presented.
\end{abstract}

@@

\par LoRa can adjust data rate by spreading symbols within a fixed channel bandwidth. This enables tradeoff between receive sensitivity and transmission time on air \cite{7803607}. Simultaneous same spreading factor (SF) transmissions are prone to collision, however, different SF transmissions in the same channel are orthogonal to each other, thus SF assignment is crucial for overall network performance. In this work, a LoRa discrete event simulator is developed to study different LoRa SF assignment strategies. We also proposed an SF assignment scheme to increase network performance with the help of machine learning methods and we verified our proposal using the simulator.

@@

\section{LoRa}
\par LoRa is the name of the proprietary physical layer radio/chipset technology that provides wireless link solution for low power wide area networks. LoRa is designed and patented by Semtech Corporation. LoRa uses proprietary spread spectrum modulation technique that is derivative of Chirp Spread Spectrum (CSS). A chirp is a sinusoidal signal which its frequency increases over time. Chirp frequency increases linearly and sweeps the entire bandwidth. We invite readers to refer to \cite{AN1200.22} for further details.

@@

\subsection{Spreading Factor Assignment Issue}
\par Simultaneous different SF transmissions are orthogonal to each other up to some extent. Which means, a LoRa gateway can simultaneously receive multiple transmissions with different SFs. However, simultaneous transmissions with same SF may not be received by gateway due to collision. For this reason, spreading factor assignment of nodes is crucial for network performance.

\par In a LoRaWAN network, initially, a node is not aware of how far it is from a GW. However, a node can guess the distance from a gateway by observing received signal power of a downlink transmission. If a downlink transmission received signal power is too high, then node can decrease its next transmission SF to decrease power consumption. This SF assignment method is called lowest possible SF assignment scheme for the rest of the paper. Also, a GW can request from a node to decrease node's SF or transmit power.

@@

\section{LoRaWAN} \label{LoRaWAN}
\par LoRa has an open standard medium access control (MAC) layer protocol called LoRaWAN. LoRaWAN is a medium access control (MAC) layer protocol which designed for large scale LoRa networks. LoRaWAN is developed and maintained by LoRa Alliance. LoRa Alliance is an open, non-profit organization dedicated to standardization of LoRaWAN. LoRaWAN  provides inter-operability between different LoRa networks. LoRaWAN is based on pure ALOHA medium access which means end nodes do not check whether the channel is free or not before transmitting, accepting the possibility of the collision.

\par A typical LoRaWAN network consist of three network entities.

\subsection{End node}
\par LoRaWAN end node (EN) is a low power embedded device that only communicates to gateways. LoRaWAN standard defines three classes for end devices which are Class A, Class B, and Class C. Different classes provides LPWAN solutions to different deployments and topologies. Class A end nodes generate uplink transmission at any time and only receives a period of time after uplink transmission. Class B end nodes extend Class A behavior by adding scheduled receive windows for downlink transmission. Receive window is synchronized using a beacon packet transmitted by gateways. Class C end nodes extend Class A behavior by keeping receive window open all the time except uplink transmission. This provides Class C end nodes to low latency downlink communication, but it requires more power consumption. In this paper, we focused on Class A end devices since Class A behavior provides the lowest power consumption.

\subsection{Gateway}
\par LoRaWAN gateway (GW) is a device that receive/transmit packets coming from/to end nodes. A typical gateway can receive from multiple channels at the same time. Gateways are typically connected to power grid, so power consumption of a gateway is insignificant in most of the deployments.

@@

\par In \cite{7996384}, the authors evaluated the performance of LoRaWAN networks in a smart city scenario. The authors proposed a link measurement and a link performance model for LoRa. The authors also proposed a SINR threshold matrix for modeling LoRa interference between simultaneous but different SF LoRa transmissions. They implemented a LoRa simulator in ns-3 to study scalability and performance of LoRaWAN networks. Their results show that LoRaWAN networks scale well as the number of nodes and gateways increases. They also show that SF assignment has great effect on LoRaWAN network performance.

\par In \cite{8090518}, another LoRaWAN ns-3 simulator is presented. Authors introduced an error model for determining range as well as interference between multiple simultaneous LoRa transmissions. Their simulator supports LoRaWAN Class A end devices, multiple gateways, both upstream and downstream confirmed messages. Their results show that allocating network parameters to end devices is hugely important for the performance of LoRaWAN networks.

@@

\par In \cite{8430542}, the authors investigated the impact of interference caused by simultaneous LoRa transmissions with same and different SFs. They derived aggregated co-SF and inter-SF interference power SIR distributions to capture the coverage distance from the gateway for modeling interference in multiple gateway scenarios. Their results show that transmission among different SFs can cause a significant impact in high-density LoRaWAN networks.

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} can be solved by forcing some of the close nodes to select higher SF even they can able to communicate with lower SF. This may prevent collisions due to the orthogonality of different SF as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, how network should assign which SF to which node becomes another issue. Increasing a node's SF should be done carefully since higher SF means higher time on air and higher time on air is more prone to collision with other higher SF transmissions. In multiple GW scenarios, this approach may increase the collisions with nodes in other GW's range. Thus, extra care should be taken for nodes in intersection area of GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes' SFs requires different rules. For this reason, we proposed a machine learning based SF assignment scheme to decrease same SF transmission collisions. This technique works by learning the behavior of the nodes in a network. A NS can keep track of successful uplink transmissions and their SFs. NS can also keep track of some of the collided transmission if header part of the packet is not interfered at location of the GW. However, NS cannot keep track of transmissions with receive power lower than the GW sensitivity. Using these obtained information, NS can train a classifier for predicting future transmission result for specific node and specific SF. Using this prediction model NS can assign the SF to nodes considering the collision probability.

\par In this work, decision tree classifier (DTC) and support vector machine (SVM) classifier are evaluated to predict transmission results. Class weights are balanced according to sample distributions for both methods. For DTC, Gini impurity criteria is used to measure the quality of a splits. For SVM, penalty parameter is set to 1, degree is set to 3 and RBF kernel is used.

\par We have been able to generate mass amount of LoRaWAN transmission logs for different topologies using the simulator. Data set size is directly proportional to simulation duration. Thus, increasing simulation duration, improves the prediction accuracy. In real world deployments, NS can keep track of transmission and it can create a classifier daily basis, then GWs can request from nodes to use their new SFs.

@@

\section{Simulation Environment} \label{Simulation Environment}
\par A discrete event simulator is implemented in Python to study effects of different SF strategies in LoRaWANs. The LoRaWAN SF simulation tool source code is available at \cite{simlorasf}. Simulation tool supports custom LoRaWAN topologies as well as randomly generated LoRaWAN topologies. Simulator can generate uniformly distributed circular shape network topology with radius (m), number of nodes and number of gateways input parameters. Global simulation input parameters are simulation duration (s), packet size (B), packet generation rate (p/s) and spreading factor assignment method. With these inputs, the simulator produces total number of generated packets, number of successfully received packets, number of interfered packets, number of under sensitivity packets, network packet delivery ratio percentage (PDR), network throughput (bps) and total transmit energy consumption (J). Simulator also produces  prediction accuracy and confusion matrix for machine learning schemes.

\par In the simulator, only LoRaWAN Class A devices are covered. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet according to Poisson interval for given packet rate input parameter. In the simulator, downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictates duty cycle limit for all devices including GWs.

@@

\par In the simulator, it is assumed that nodes always select maximum allowed transmit power, which is 14 dBm for European ISM band regulation. Different channel transmissions are independent from each other. But, in this work we focus on SF orthogonality. Thus, we only utilized single channel transmissions in the simulator.

\par Receive sensitivity of a LoRa gateway for different SFs and BWs in dBm unit can be found in \cite{SX1276}. In the simulator, 125 kHz bandwidth receive sensitivities are used.

@@

\par If the received signal power is higher than the gateway sensitivity, then signal can be decoded by the receiver successfully when there is no interfering transmission at the duration of transmission.

@@

\section{Simulation Results}  \label{Simulation Results}
\par For simulation results in this paper, global simulation parameters are set as follows. Packet size is 60 bytes including header and payload, simulation duration is 3600 seconds, packet generation rate is 0.01 packet per second.

\par In Figure \ref{fig:sf_pdr}, PDR plot of different SF assignment is shown. Randomly generated network topology radius is set to 3000 meters and number of GWs is set to 1. Increasing SFs increases time on air. This increases the number of collisions thus decreases the PDR of the network. PDR of higher SFs are decreased to almost 0 while number of nodes is increasing. Since network topology radius is quite small almost all SFs can reach to the GW.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{gw_pdr_r3000_p0.01_s3600}.png}
\caption{PDR of different number of gateways.}
\label{fig:gw_pdr}
\end{figure}

@@

\par For space constraints, we omit figures for network throughput output, total transmit energy consumption output and different packet generation rate input. We invite readers to experiment the simulation tool with different inputs and outputs. \cite{simlorasf}.

\par In Figure \ref{fig:prediction_pdr}, result of proposed technique is presented. PDR plot of lowest SF scheme and machine learning prediction schemes are shown in same figure. Randomly generated network topology radius is set to 5000 meters and number of GWs is set to 3. Prediction model needs nodes location and three GWs are enough to roughly locate nodes positions by triangulation. In Table \ref{table:prediction_pdr}, PDR values for different topology radius are presented. Both prediction SF schemes give better PDR than lowest SF schemes when number of nodes increases since number of interference also increases. Prediction schemes can only improve network performance when LoRa interference is present. Also, prediction schemes give better results when nodes are deployed close to the gateway, since nodes have margin to increase their SFs. If a node is far away from the GW, then prediction schemes cannot increase the SF to avoid interference since the SF is already high.

In Table \ref{table:prediction_accuracy}, prediction accuracy of SVM and DTC prediction schemes are presented for different network topology radius and number of nodes. Number of GW is set to 3.

@@

\section{Conclusion} \label{Conclusion}
\par In this paper, after a brief introduction about LPWAN technologies, we provide background information about one of the most promising LPWAN technology, LoRa. Then, we discuss about LoRa modulation basics and spreading factor assignment issue. We have implemented a discrete event simulator from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. We have shown how same SF transmission collisions can be avoided and we proposed a machine learning based solution for collision with same SF transmissions. We illustrate simulation results for proposed machine learning SF assignment technique. Our simulation results show that, proposed technique can increase network performance for LoRaWAN networks when nodes are deployed close to the GWs.

>>>>

\begin{abstract}
Low power wide area network (LPWAN) technologies offer affordable connectivity to massive number of low-power devices distributed over large geographical areas. Focus of this work is one of the most promising LPWAN technologies: LoRa. LoRa offers long range communication and strong resilience to interference by proprietary modulation technique based on Chirp Spread Spectrum (CSS). LoRa modulation trades data rate for sensitivity and communication range by spreading symbols within a fixed channel bandwidth. Collisions in LoRaWAN networks are strongly related with spreading factor (SF) assignment of nodes which indeed effects network performance. In this work, a simulation environment to study different SF assignment schemes is implemented. Furthermore, a novel SF assignment strategy is proposed which utilizes a machine learning technique for optimization of SF assignment. Finally, promising simulation results for proposed machine learning SF assignment technique is presented.
\end{abstract}

@@

\par LoRa can adjust data rate by spreading symbols within a fixed channel bandwidth. This enables tradeoff between receive sensitivity and transmission time on air \cite{7803607}. Simultaneous same SF transmissions are prone to collision, however, different SF transmissions in the same channel are orthogonal to each other, thus SF assignment is crucial for overall network performance. In this work, a LoRa discrete event simulator is developed to study different LoRa SF assignment strategies. We also proposed an SF assignment scheme to increase network performance with the help of machine learning methods and we verified our proposal using the simulator.

@@

\section{LoRa}
\par LoRa is the name of the proprietary physical layer radio/chipset technology that provides wireless link solution for LPWANs. LoRa is designed and patented by Semtech Corporation. LoRa uses proprietary spread spectrum modulation technique that is derivative of Chirp Spread Spectrum (CSS). A chirp is a sinusoidal signal which its frequency increases over time. Chirp frequency increases linearly and sweeps the entire bandwidth. We invite readers to refer to \cite{AN1200.22} for further details.

@@

\subsection{Spreading Factor Assignment Issue}
\par Simultaneous different SF transmissions are orthogonal to each other up to some extent. Which means, a LoRa GW can simultaneously receive multiple transmissions with different SFs. However, simultaneous transmissions with same SF may not be received by GW due to collision. For this reason, SF assignment of nodes is crucial for network performance.

\par In a LoRaWAN network, initially, a node is not aware of how far it is from a GW. However, a node can guess the distance from a GW by observing received signal power of a downlink transmission. If a downlink transmission received signal power is too high, then node can decrease its next transmission SF to decrease power consumption. This SF assignment method is called lowest possible SF assignment scheme for the rest of the paper. Lowest possible SF assignment scheme is commonly used in LoRaWAN deployments. GW can request from a node to decrease its SF or transmit power.

@@

\section{LoRaWAN} \label{LoRaWAN}
\par LoRa has an open standard medium access control (MAC) layer protocol called LoRaWAN which designed for large scale LoRa networks considering well known LPWAN challenges and their best practice solutions. LoRaWAN is developed and maintained by LoRa Alliance. LoRa Alliance is an open, non-profit organization dedicated to standardization of LoRaWAN. LoRaWAN is based on pure ALOHA medium access which means end nodes do not check whether the channel is free or not before transmission, accepting the possibility of a collision. A typical LoRaWAN network consist of following three network entities.

\subsection{End node}
\par LoRaWAN end node (EN) is a low power embedded device that only communicates to GWs. LoRaWAN standard defines three classes for end devices which are Class A, Class B, and Class C. Different classes provides LPWAN solutions to different applications and deployments. Class A end nodes generate uplink transmission at any time and only receives a period of time after uplink transmission. Class B end nodes extend Class A behavior by adding scheduled receive windows for downlink transmission. Receive window is synchronized using a beacon packet transmitted by GWs. Class C end nodes extend Class A behavior by keeping receive window open all the time except uplink transmission. This provides Class C end nodes to low latency downlink communication, but it requires more power consumption. In this paper, we focused on Class A end devices since Class A behavior provides the lowest power consumption.

\subsection{Gateway}
\par LoRaWAN gateway (GW) is a device that communicates to end nodes. A typical GW can receive from multiple channels at the same time. GWs are typically connected to power grid, so power consumption of a GW is insignificant in most of the deployments.

@@

\par In \cite{7996384}, the authors evaluated the performance of LoRaWAN networks in a smart city scenario. The authors proposed a link measurement and a link performance model for LoRa. The authors also proposed a SINR threshold matrix for modeling LoRa interference between simultaneous but different SF LoRa transmissions. They implemented a LoRa simulator in ns-3 to study scalability and performance of LoRaWAN networks. Their results show that SF assignment has great effect on LoRaWAN network performance.

\par In \cite{8090518}, another LoRaWAN ns-3 simulator is presented. Authors introduced an error model for determining range as well as interference between multiple simultaneous LoRa transmissions. Their simulator supports LoRaWAN Class A end devices, multiple GWs, both upstream and downstream confirmed messages. Their results show that allocating network parameters to end devices is hugely important for the performance of LoRaWAN networks.

@@

\par In \cite{8430542}, the authors investigated the impact of interference caused by simultaneous LoRa transmissions with same and different SFs. They derived aggregated co-SF and inter-SF interference power SIR distributions to capture the coverage distance from the GW for modeling interference in multiple GWs scenarios. Their results show that transmission among different SFs can cause a significant impact in high-density LoRaWAN networks.

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} can be solved by forcing some of the close nodes to select higher SF even they can able to communicate with lower SF. This may prevent collisions due to the orthogonality of different SF as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, how network should assign which SF to which node becomes an important question. Increasing a node's SF should be done carefully since higher SF means higher time on air and higher time on air is more prone to collision with other high SF transmissions. In multiple GW scenarios, this approach may increase the collisions with nodes in other GW's range. Thus, extra care should be taken for nodes in intersection area of GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes' SFs requires different rules. For this reason, we proposed a machine learning based SF assignment approach to decrease same SF transmission collisions. This technique works by learning the transmission behavior of the nodes in a network. A NS can keep track of successful uplink transmissions and their SFs. NS can also keep track of some of the collided transmission if header part of the packet is not interfered at location of the GW. However, NS cannot keep track of transmissions with receive power lower than the GW sensitivity. Using these obtained information, NS can train a classifier for predicting future transmission result for specific node and specific SF. Using this prediction model NS can assign the SF to nodes considering the collision probability.

\par In this work, decision tree classifier (DTC) and support vector machine (SVM) schemes are evaluated to predict transmission results. Class weights are balanced according to sample distributions for both methods. For DTC, Gini impurity criteria is used to measure the quality of a splits. For SVM, penalty parameter is set to 1, degree is set to 3 and RBF kernel is used.

\par We have been able to generate mass amount of LoRaWAN transmission records for different topologies using the simulator. Data set size is directly proportional to simulation duration. Thus, increasing simulation duration, improves the prediction accuracy. In real world deployments, NS can keep track of all the transmission and it can create a classifier daily basis, then GWs can request from nodes to use their new SFs.

@@

\section{Simulation Environment} \label{Simulation Environment}
\par A discrete event simulator is implemented in Python to study effects of different SF strategies in LoRaWANs. LoRaWAN SF simulation tool source code is available at \cite{simlorasf}. Simulation tool supports custom LoRaWAN topologies as well as randomly generated LoRaWAN topologies. Simulator can generate uniformly distributed circular shape network topology with input parameters; radius (m), number of nodes and number of GWs. Global simulation input parameters are simulation duration (s), packet size (B), packet generation rate (p/s) and SF assignment method. With these inputs, the simulator produces total number of generated packets, number of successfully received packets, number of interfered packets, number of under sensitivity packets, network packet delivery ratio percentage (PDR), network throughput (bps) and total transmit energy consumption (J). Simulator also produces prediction accuracy and confusion matrix for machine learning schemes.

\par In the simulator, only LoRaWAN Class A devices are covered. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet randomly according to Poisson interval for given packet rate input parameter. In the simulator, downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictates duty cycle limit for all devices including GWs.

@@

\par In the simulator, it is assumed that nodes always select maximum allowed transmit power, which is 14 dBm for European ISM band regulation \cite{lorawan.regional.parameters}. Different channel transmissions are independent from each other. But, in this work we focus on SF orthogonality. Thus, we only utilized single channel transmissions in the simulator.

\par Receive sensitivity of a LoRa GW for different SFs and BWs in dBm unit can be found in \cite{SX1276}. In the simulator, 125 kHz bandwidth receive sensitivities are used.

@@

\par If the received signal power is higher than the GW sensitivity, then signal can be decoded by the receiver successfully when there is no interfering transmission.

@@

\section{Simulation Results}  \label{Simulation Results}
\par For simulation results in this paper, global simulation parameters are set as follows; packet size is 60 bytes including header and payload, simulation duration is 3600 seconds, packet generation rate is 0.01 packet per second.

\par In Figure \ref{fig:sf_pdr}, PDR plot of different SF assignment is shown. Randomly generated network topology radius is set to 3000 meters and number of GWs is set to 1. Increasing SFs, increases time on air. This increases the number of collisions thus decreases the PDR. Higher SFs effected exponentially while number of nodes increases. Since network topology radius is quite short, all SFs can reach to the GW.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{gw_pdr_r3000_p0.01_s3600}.png}
\caption{PDR of different number of GWs.}
\label{fig:gw_pdr}
\end{figure}

@@

\par For space constraints, we omit figures for network throughput output, total transmit energy consumption output and different packet generation rate input. We invite readers to experiment the simulation tool with different input parameters and their results. \cite{simlorasf}.

\par In Figure \ref{fig:prediction_pdr}, result of proposed technique is presented. PDR plot of lowest SF scheme and machine learning prediction schemes are shown in same figure. Randomly generated network topology radius is set to 5000 meters and number of GWs is set to 3. Prediction model needs nodes location and three GWs are enough to roughly locate nodes positions by triangulation. In Table \ref{table:prediction_pdr}, PDR values for different topology radius are presented. Both prediction SVM and DTC SF schemes give better PDR than lowest SF schemes when number of nodes increases. Increasing number of nodes, increases number of interference. Prediction schemes  improve network performance when LoRa interference is high. Also, prediction schemes give better results when nodes are deployed close to the GW, since nodes have margin to increase their SFs when deployed close to the GW. If a node is far away from the GW, then prediction schemes cannot increase the SF to avoid interference since the SF is already high.

\par In Table \ref{table:prediction_accuracy}, prediction accuracy of SVM and DTC schemes are presented for different network topology radius and number of nodes. Prediction accuracy is not directly proportional to network PDR. Correct prediction of a interfered transmission may not increase the PDR but increases the accuracy metric. DTC gives better network PDR results than SVM, even overall SVM prediction accuracy is high. LoRaWAN simulation usually produces imbalanced dataset, DTC gives better accuracy for interfered transmission labels. We invite readers to experiment the simulation tool to review confusion matrices for different machine learning schemes.

@@

\section{Conclusion} \label{Conclusion}
\par In this paper, after a brief introduction about LPWAN technologies, we provide background information about LoRa and LoRaWAN. Then, we discuss about LoRa modulation basics and spreading factor assignment issue. We have implemented a discrete event simulator from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. We have shown how same SF transmission collisions can be avoided and we proposed machine learning based solution for collision with same SF transmissions. We illustrate simulation results for proposed machine learning SF assignment technique. Our simulation results show that, proposed technique can increase network performance for LoRaWAN networks especially nodes are deployed close to the GWs.
